---
author: "Steven Ren"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, fig.height = 4)
library(tidyverse)
library(lubridate)
library(scales)
library(modelr)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

\renewcommand{\prob}{\mathsf{P}}
\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\SD}{\mathsf{SD}}
\newcommand{\SE}{\mathsf{SE}}

## Homework Assignment 11

#### Due Friday, April 28, 2022, at 11:59 PM

### Preliminaries

- Directories
    - COURSE/homework/
    - COURSE/homework/hw11/
    - COURSE/data/
    - COURSE/scripts/
- Files
  - COURSE/homework/hw11/hw11.Rmd
  - COURSE/data/boston-marathon-data.csv
  - COURSE/data/dugong.csv
  - COURSE/scripts/viridis.R
  - COURSE/scripts/ggprob.R

### Data

- Some problems use the official Madison weather data, `madison-weather-official-1869-2022.csv`.
- Additional problems use the Boston Marathon data in the file `boston-marathon-data.csv`. This file is a transformed version of the raw data we used in class and has data for all runners who completed the race in 2010 and 2011. The variable `Time` is the sum of the times from the different portions of the race, each of which begins with "K".

### Aims

- Practice regression

## Problems

  **1.** In a regression problem to estimate $y$ from explanatory variable $x$ from a sample of size $n$, partial summary information is $\bar{x} = 20$ and $\bar{y} = 100$. Regardless of the values of other summary statistics, what is the value the predicted value $\hat{y}$ at a point where $x = 20$? Briefly explain.
  
> The predicted value of y_hat is 100.



  **2.** In a regression problem to estimate $y$ from explanatory variable $x$ from a sample of size $n$, partial summary information is $\bar{x} = 20$, $s_x = 5$, $\bar{y} = 100$, and $s_y = 15$. Which of the following values are possible values for the predicted value $\hat{y}$ when the explanatory variable has value $x = 30$? Briefly explain.
  
**(a)** 50      
**(b)** 70      
**(c)** 100      
**(d)** 120    
**(e)** 150

```{r}
n = 100
x = rnorm(n, mean = 20, sd = 5)
y = rnorm(n, mean = 100, sd = 15)

r = cor(x, y)

x0 = 30
z = (x0 - 20)/5
z

yhat = 100 + r*z*15
yhat
```

> The predicted y_hat value is option (c) 100 after using a sample linear regression



Problems 3--6 are based on the data set in the file *dugong.csv* which relates age (in years) and length (in meters) of a sample of 27 dugongs, a type of marine mammal.
  
Credit:  The *dugong.csv* file is from Data8 at UC-Berkeley.

  **3.**

- Read in the *dugong.csv* data set.  
-  Create a scatter plot with `length` on the x-axis and `age` on the y-axis; be sure to add descriptive axis labels (include units of measurement) and a title.  
-  Using `geom_smooth()`, add the least-squares line to your plot.

```{r}
dugong = read_csv("../../data/dugong.csv")

ggplot(dugong, aes(x = Length, y = Age)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Length (m)", y = "Age (years)", title = "Scatter Plot of Dugong Length vs Age")
```



  **4.**

- Using the dugong data, calculate the sample means, sample standard deviations, and correlation coefficient of the variables `age` and `length`.
- Using formulas from lecture, calculate the slope and intercept of the least squares regressions line to predict age with length.

```{r}
mean_age = mean(dugong$Age)
mean_length = mean(dugong$Length)
mean_age
mean_length

sd_age = sd(dugong$Age)
sd_length = sd(dugong$Length)
sd_age
sd_length

r = cor(dugong$Age, dugong$Length)

slope = r * (sd_age / sd_length)
intercept = mean_age - slope * mean_length
slope
intercept
```

- Use the dugong data and the functions `lm()` and `coef()` to calculate the slope and intercept of the least squares regression line of age against length (use length to predict age).

```{r}
model = lm(Age ~ Length, data = dugong)
intercept = coef(model)[1]
slope = coef(model)[2]
slope
intercept
```

- Verify that you get the same values.

> Both result in a slope of 23.77168 and intercept of -44.56683.



  **5.**

- Add columns with the predicted values and residuals to the dugong data set. *(You can use* **modelr** *functions or just use `mutate()` and calculate these values directly.)*
- Plot the residuals versus length.
    - Add a horizontal line at $y=0$ and appropriate labels on each axis.

```{r}
dugong = dugong %>%
  mutate(Predicted = predict(model), Residual = Age - Predicted)

ggplot(dugong, aes(x = Length, y = Residual)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Length (cm)", y = "Residuals")
```

- Describe what the residual plot suggests about the appropriateness of using simple linear regression to predict age from length of dugongs.

> The residual plot suggets that the using simple linear regression to predict age from length of dugongs is not appropriate, because a lot of residuals are far off from 0, a good residual plot should have msot values close to 0.



  **6.**

- Print the summary of the fitted regression model

```{r}
summary(dugong)
```

- The simple linear regression model for $Y_i$ conditional on the values of $X_i = x_i$ is

$$
\E(Y_i \mid X_i = x_i) = \beta_0 + \beta_1 x_i + \varepsilon_i, \quad \text{for $i = 1, \ldots,n$}
$$

where $\varepsilon_i \sim \text{Normal}(0, \sigma)$
for some parameter $\sigma > 0$.

- The parameter $\sigma$ is the unknown population standard deviation of the typical distance between a point $Y_i$ and its true expected value.

- Use the function `sigma()` on the fitted regression object (what you created with `lm()`) to extract the estimate of $\sigma$. Identify where this numerical value appears in the printed summary you made earlier.

```{r}
sigma = sigma(model)
sigma
```
sigma is the residual's standard error

- The numerical estimate of $\sigma$ here is not quite the standard deviation of the residuals because the denominator is $n-2$, the degrees of freedom in simple linear regression, instead of $n-1$, the degrees of freedom from a single numerical sample.

- Use the column of residuals in the augments data set `dugong` and verify that:
    - the mean of the residuals equals zero (numerically, it might be very close).
    - you arrive at the numerical estimate of $\sigma$ by calculating
    
$$
\sqrt{ \frac{\sum_{i=1}^n (y_i - \hat{y}_i)^2}{n-2} }
$$

where the $i$th residual is $y_i - \hat{y}_i$.

```{r}
mean_residual = mean(dugong$Residual)
mean_residual
sigma = sqrt(sum(dugong$Residual^2) / (nrow(dugong) - 2))
sigma
```



- Problems 7--8 use the cleaned Boston Marathon data in `boston-marathon-data.csv`.

  **7.**

- Read in the Boston marathon data from the file `boston-marathon-data.csv`.

```{r}
boston = read_csv("../../data/boston-marathon-data.csv")
```

- Create a scatter plots of `Time` versus `Age` for the female runners in 2010.
    - Add a straight regression line
    - Add a smooth curve
- As there are so many points, you may set `alpha` to a value less than one inside of `geom_point()` to lessen the effects of overplotting.    
    
```{r}
female_2010 = boston %>% 
  filter(Year == 2010, Sex == "female")

ggplot(female_2010, aes(x = Age, y = Time)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Age", y = "Time", 
       title = "Time vs Age for Female Runners in 2010")
```
    
- Make a residual plot of the residuals versus `Age`.
    - Include a horizontal line at $y=0$
    - Include a smooth curve through the residuals

- In addition, make a density plot of the residuals.    
```{r}
model = lm(Time ~ Age, data = female_2010)
female_2010_residual = female_2010 %>%
  mutate(Predicted = predict(model), Residual = Time - Predicted)

ggplot(female_2010_residual, aes(x = Age, y = Residual)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_hline(yintercept = 0) +
  labs(x = "Age", y = "Residuals") +
  ggtitle("Time vs. Age for Female Runners in 2010 with Residuals")

ggplot(female_2010_residual, aes(x = Residual)) +
  geom_density(alpha = 0.5) +
  labs(x = "Residuals", y = "Density") +
  ggtitle("Time vs. Age for Female Runners in 2010 Density Plot")
```



  **8.** Examine the residual plots from the previous problem.
  
- Is there evidence of strong non-linearity?

> No there is not, the residual graph has a horizontally sloped line at y = 0

- Is there evidence that the standard deviation of the residuals varies substantially with changes in age?

> Yes, it looks like the stadard deviation decreases as age increases based off the residual plot.


- Is there evidence that the error distribution for individual residuals is not symmetric?

> Yes there is some evidence, the density plot seems to be left skewed.


