---
author: "Steven Ren"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE, error = TRUE, fig.height = 3)
library(tidyverse)
library(kableExtra)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}
\renewcommand{\prob}{\mathsf{P}}


## Assignment 7

#### Due Sunday, April 2, 11:59 PM CT

### Preliminaries

- Directories
    - COURSE/homework/
    - COURSE/homework/hw07/
    - COURSE/data/
    - COURSE/scripts/
- Files
  - COURSE/homework/hw07/hw07.Rmd
  - COURSE/scripts/viridis.R
  - COURSE/scripts/ggprob.R

### Aims

- Practice probability and the binomial distribution

### Problems

  1. A discrete random variable $X$ has possible values 0, 1, 2, 3, 4, 5, 6, 7, 8 with the following partial probability distribution.
The missing probabilities $\prob(X=7)$ and $\prob(X=8)$ are equal to each other.
```{r, echo = FALSE}
set.seed(20221021)
x = 0:8
p = rnorm(7,5,2.1)
p = sort(round(p / sum(p) * 0.84, 2))
p1 = tibble(x = x, p = c(p, rep(NA,2)))

p1_tab = p1 %>% 
  mutate(x = str_c(x)) %>% 
  pivot_wider(names_from = x, values_from = p) %>% 
  mutate(x = "P(X=x)", .before = `0`) 

p1_tab %>% 
  kable() %>% 
  kable_styling(position = "center", full_width = FALSE)
```

- Put the values `x = 0:8` and the tabled probabilities `p` into a data frame (use `tibble()`).
```{r}
table1 = tibble(x = x, p = p1$p)

table1
```

- Write code to calculate the missing values for $\prob(X=7) = \prob(X = 8)$ and replace the missing values with these calculated values into your data frame.
```{r}
sum = sum(table1$p[1:7])
sum_missing = 1 - sum
missing = sum_missing / 2
table1$p[8:9] = missing

table1
```

- Verify that the sum of all probabilities is 1.
```{r}
sum = sum(table1$p[1:9])

sum
```

- Calculate the mean, variance, and standard deviation of this distribution.
```{r}
mu = sum(table1$x * table1$p)
mu

variance = sum((table1$x^2 * table1$p)) - mu^2
variance

sd = sqrt(variance)
sd
```



  2. Suppose you have a random variable $X \sim \text{Binomial}(120, 0.2)$.
  
- 2a. Calculate and report the mean and standard deviation of the distribution.
```{r}
mean = 120 * 0.2
mean

variance = 120 * 0.2 * (1 - 0.2)
variance

sd = sqrt(variance)
sd
```

- 2b. Calculate and report the probability that the random variable is exactly equal to 20, $\prob(X = 20)$.
```{r}
dbinom(20, 120, 0.2)
```

- 2c. Calculate and report the probability that $X$ equals 20 or more, $\prob(X \ge 20)$.
```{r}
1 - pbinom(19, 120, 0.2)
```

- 2d. Draw a graph of the binomial distribution with vertical lines at the mean (red and dashed), and 1, 2, and 3 standard deviations above and below the mean (red and dotted).
```{r}
gbinom(n=120, p=0.2, size=3) +
  geom_bar(stat = "identity", fill = "blue", color = "black") +
  geom_vline(xintercept = mean, color = "red", linetype = "dashed") +
  geom_vline(xintercept = mean - sd, color = "red", linetype = "dotted") +
  geom_vline(xintercept = mean + sd, color = "red", linetype = "dotted") +
  geom_vline(xintercept = mean - 2*sd, color = "red", linetype = "dotted") +
  geom_vline(xintercept = mean + 2*sd, color = "red", linetype = "dotted") +
  geom_vline(xintercept = mean - 3*sd, color = "red", linetype = "dotted") +
  geom_vline(xintercept = mean + 3*sd, color = "red", linetype = "dotted") +
  xlab("X") + 
  ylab("Probability") +
  ggtitle("Binomial Distribution of X")
```



  3. What is the probability that $X$ from Problem 2 is within one, two, and three standard deviations of the mean? Round each probability to four decimal places.
```{r}
within1sd = round((pbinom(mu + sd, 120, 0.2) - pbinom(mu - sd, 120, 0.2)), 4)

within2sd = round((pbinom(mu + 2 * sd, 120, 0.2) - pbinom(mu - 2 * sd, 120, 0.2)), 4)

within3sd = round((pbinom(mu + 3 * sd, 120, 0.2) - pbinom(mu - 3 * sd, 120, 0.2)), 4)

within1sd
within2sd
within3sd
```



  4. The following code makes a graph of the cumulative distribution function for the $\text{Binomial}(7, 0.3)$ distribution, that is $F(x) = \prob(X \le x)$. This is the function which is calculated by the base R function `pbinom()`.

- Modify the code to add a horizontal red dotted line at 0.4.
```{r}
prob4 = tibble(
  x = seq(-1,8,1),
  p = dbinom(x, 7, 0.3),
  cdf = pbinom(x, 7, 0.3))

ggplot(prob4, aes(x = x, y = cdf )) +
  geom_step(color = "blue") +
  geom_hline(yintercept = 0) +
  scale_x_continuous(breaks = 0:7) +
  geom_hline(yintercept = 0.4, color = "red", linetype = "dotted") +
  xlab("x") +
  ylab("Probability") +
  ggtitle("Binomial(7, 0.3) distribution CDF") +
  theme_minimal()
```

- Calculate the 0.4 quantile of the $\text{Binomial}(7, 0.3)$
```{r}
qbinom(0.4, 7, 0.3)
```

- Explain how the value of the quantile is related to the red dotted line which you added to the plot.

> The red dotted line of prob = 0.4 intersects at x = 2

- Use the graph to determine the 0.75 quantile of this distribution.
Then verify your observation with a calculation.

```{r}
qbinom(0.75, 7, 0.3)
```

> 3



  5. In a large population of likely voters, the proportion of voters who plan to vote for candidate A is $p = 0.52$. In a random sample of 600 likely voters, what is the probability that the proportion of voters who plan to vote for candidate A is less than 50% (the number of such voters is less than 300)?
  
```{r}
pbinom(299, 600, 0.52)
```



  6. Basketball player 1 makes 80% of their free throws and basketball players 2 makes 70% of their free throws. Assume all of their free throws attempts are independent. If each player takes ten free throws and $X$ is the total number of free throws they make, does $X$ have a binomial distribution? If yes, specify the parameters $n$ and $p$. If not, explain why not.
  
> Yes x has a binomial distribution, n = 10 + 10 = 20, p = (10*0.7 + 10*0.8)/20 = 0.75



  7. What is the smallest number of coin tosses $n$ of a fair coin so that the chance of getting 50 or more heads exceeds 90%?  (You may need to do some trial and error. Alternatively, think of creating a data frame with a column `n` and a column for the probability and using this to find the answer.)

```{r}
1 - pbinom(49, 113, 0.5)
```

> The smallest number of coin tosses n is 113 tosses



  8. The random variable $X$ has the $\text{Binomial}(100, 0.2)$ distribution.
  
- Find an integer $a$ so that $P(X \le a) \ge 0.5$ and $P(X \ge a) \ge 0.5$.
- Show the values of $a$, $P(X \le a)$, and $P(X \ge a)$.

```{r}
mean = 100 * 0.2
pbinom(20, 100, 0.2)
1-pbinom(19, 100, 0.2)
```

> a = 20, P(X≤a) = 0.5594616, and P(X≥a) = 0.5398386
  


